A general model for the evolution of nuptial gift-giving

### 2024-11-29 index.Rmd
Fixed up traits and manuscript. Traits are not binary anymore, but are diploid

---
title: "A general model for the evolution of nuptial gift-giving"
author: |
  | Anders P. Charmouh$^{1,4, 5}$, Trine Bilde$^{2}$, Greta Bocedi$^{1}$, A. Bradley Duthie$^{3}$
date: '[1] School of Biological Sciences, University of Aberdeen, UK. [2] Department of Biology, Aarhus University, Denmark. [3] Biological and Environmental Sciences, University of Stirling, UK.  [4] Bioinformatics Research Centre Aarhus University, University City 81, building 1872, 3rd floor, DK-8000 Aarhus C, Denmark [5] Email: apc@birc.au.dk'
output:
  html_document: default
  pdf_document:
    fig_caption: yes
    keep_tex: yes
  word_document:
    fig_caption: yes
    pandoc_args:
    - --csl
    - evolution-letters.csl
    reference_docx: docx_template.docx
bibliography: nuptial_refs.bib
header-includes:
- \usepackage{amsmath}
- \usepackage{natbib}
- \usepackage{lineno}
- \usepackage{caption}
- \usepackage[utf8]{inputenc}
- \bibliographystyle{amnatnat}
linestretch: 1
link-citations: yes
linkcolor: blue
csl: evolution-letters.csl
biblio-style: apalike
---

```{r, echo = FALSE}
library("knitr")
calc_beta <- function(M = 1, TF = 1, TM = 1){
  term1 <-   (M * (TF - TM))^2;
  term2 <-   (M * (TF - TM))^4;
  term3 <- 4*(M * (TF - TM))^2;
  ttrm  <- (term2 + term3)^(1/2);
  sol1  <- 0.5 * (term1 - ttrm) + 1;
  sol2  <- 0.5 * (term1 + ttrm) + 1;
  if(TF > TM){
    beta = max(c(sol1, sol2));
  }else{
    beta = min(c(sol1, sol2));
  }
  return(list = c(beta = beta, solution1 = sol1, solution2 = sol2));
}
get_beta <- function(M = 1, TF = 1, TM = 1){
  betas <- rep(NA, length = length(TM));
  for(i in 1:length(TM)){
    betas[i] <- calc_beta(M = M, TF = TF, TM = TM[i])[[1]];
  }
  return(betas);
}
PrGf <- function(Tm, a1){
  vals <- 1 - exp(-(1/a1)*Tm);
  return(vals);
}
PrFf <- function(Tm, a1, a2 = Inf){
  vals <- (1 - exp(-(1/a2)*Tm)) * exp(-(1/a1)*Tm);
  return(vals);
}
dWdT <- function(gamma, Tm, beta = 1, M = 100, a1 = 1){
  num1 <- gamma * (M*M + ((M*M*Tm + M*sqrt(beta))/a1) );
  dem1 <- exp((1/a1)*Tm);
  num2 <- (num1 / dem1) - M*M*(1 + gamma);
  den2 <- ((Tm * M) + sqrt(beta))^2;
  val  <- num2/den2;
  return(val);
}
dWdT_check <- function(gamma, beta = 1, M = 100, a1 = 1){
  term1 <- (gamma * M) / (a1*sqrt(beta));
  term2 <- (M*M) / beta;
  val   <- term1 - term2;
  return(val);
}
recursive_b <- function(B, D, Tf, Tm, crit = 0.0001, maxit = 9999){
  conv <- 1;
  iter <- 0;
  while(conv > crit & iter < maxit){
    Fe   <- D * (B / (1 + B));
    Me   <- (1 + B) / D;
    Bn   <- (Me / (Tm + Me)) / (Fe / (Tf + Fe));
    iter <- iter + 1;
    conv <- abs(Bn - B);
    B    <- Bn;
  }
  return(list(B = B, conv = conv, iter = iter));
}
calc_beta <- function(M = 1, TF = 1, TM = 1){
  vals <- recursive_b(B = 1, D = M, Tf = TF, Tm = TM);
  beta <- vals$B
  sol1 <- vals$conv;
  sol2 <- vals$iter;
  return(list = c(beta = beta, solution1 = sol1, solution2 = sol2));
}
get_beta <- function(M = 1, TF = 1, TM = 1){
  betas <- rep(NA, length = length(TM));
  for(i in 1:length(TM)){
    betas[i] <- calc_beta(M = M, TF = TF, TM = TM[i])[[1]];
  }
  return(betas);
}
gamma_m <- function(R = 1, beta = 1, alpha = 1/16, TF = 2){
  return(alpha * (R / (beta + 1)));
}
gamma_f <- function(R = 1, beta = 1, alpha = 1/16, TF = 2){
  topp <- exp(alpha / alpha);
  bott <- TF * R * (beta / (beta + 1)) * (exp(alpha/alpha) - 1);
  return(topp/bott);
}
PrGf <- function(Tm, a1){
  vals <- 1 - exp(-(1/a1)*Tm);
  return(vals);
}

@@

Nuptial gift-giving occurs in several taxonomic groups including insects, snails, birds, squid, arachnids and humans. 
Although this trait has evolved many times independently, no general framework has been developed to predict the conditions necessary for nuptial gift-giving to evolve. 
We use a time-in time-out model to derive analytical results describing the requirements necessary for selection to favour nuptial gift-giving. 
Specifically, selection will favour nuptial gift-giving if the fitness increase caused by gift-giving exceeds the product of expected gift search time and encounter rate of the opposite sex. 
Selection will favour choosiness in the opposite sex if the value of a nuptial gift exceeds the inverse of the time taken to produce offspring multiplied by the rate at which mates with nuptial gifts are encountered. 
Importantly, selection can differ between the sexes, potentially causing sexual conflict. 
We test these results using an individual-based model applied to a system of nuptial gift-giving spiders, *Pisaura mirabilis*, by estimating parameter values using experimental data from several studies. 
Our results provide a general framework for understanding when the evolution of nuptial gift-giving can occur and provide novel insight into the evolution of worthless nuptial gifts, occurring in multiple taxonomic groups with implications for understanding parental investment.

@@

Nuptial gift-giving occurs when the choosy sex (usually the female) receives gifts from the opposite sex (usually the male) during courtship. 
It is a widespread phenomenon, occurring within several diverse taxonomic groups such as insects, snails, birds, squid, arachnids and humans [@Lewis2014; @Lewis2012; @Albo2014]. 
Despite the ubiquity of this behaviour, little effort has been made to conceptualise the evolution of nuptial gift-giving within a general modelling framework [@Lewis2014; @Iwasa2022]. 
Recent models describing the evolution of nuptial gift-giving have focused on co-evolution between male nuptial gift-giving and female propensity to remate, and evolutionarily stable nuptial gift sizes [@Iwasa2022; @Kamimura2021], but a general framework describing the conditions necessary for nuptial gift-giving to be initially favoured by selection is needed to understand when gift-giving should evolve.

Nuptial gift-giving may allow males to increase fitness by acquiring additional mates, indirect benefits (by increasing the fitness of offspring), prolonged copulations, and success in sperm competition [@Lewis2014; @Albo2013; @Ghislandi2014]. 
However, this potential fitness increase comes at the expense of producing a nuptial gift, which may be costly in terms of time and resources. 
Females may increase their fitness by receiving nutritionally valuable nuptial gifts, but expressing a preference for males with gifts might result in a mating opportunity cost if available males without gifts are rejected. 
With respect to nuptial gift-giving, the evolutionary interests of both sexes may not always fully overlap. 
This can cause sexual conflict, which is a difference in the evolutionary interest between sexes that occurs when interaction between sexes results in a situation where neither sex can achieve an optimal outcome [@Parker2006].
That is, under some conditions, it might for example be optimal for males but not females to mate if females do not benefit from mating with males without nuptial gifts.

Much work has sought to explain how gift-giving tactics are maintained, with explanations including condition-dependent strategies, gift-giving as a way to decrease female aggression during copulation, or gifts as sensory traps [@Albo2019; @Lubin2007; @Ghislandi2018; @Toft2016]. 
An example of such a system is the nuptial gift-giving nursery-web spider *Pisaura mirabilis* where males may court females with or without nuptial gifts [@Bristowe1926; @Tuni2013a]. 
Here, males may provide females with costly nuptial gifts in the form of captured arthropod prey, and females may exhibit preference for males with a nuptial gift by rejecting males without a nuptial gift [@Albo2013].

We develop a general framework for investigating the evolution of nuptial gift-giving and choosiness using a time-in, time-out modelling approach and an individual-based model [@Clutton-Brock1992]. 
Specifically, we derive conditions under which selection will favour male search for nuptial gifts and female rejection of gift-less males. 
We show that selection for searching and choosiness depends on whether a threshold fitness value of the nuptial gift is exceeded. 
Our model demonstrates the importance of nuptial gift cost, sex ratio, and mate encounter rate in determining the threshold above which selection will favour the evolution of nuptial gift-giving.
Importantly, we show that the threshold value differs for males and females. 
We test predictions of our analytical model by formulating an individual-based model, which further supports the main theoretical results of our analytical model. 
Further, we apply our model to an example system with nuptial gifts, the nursery web spider *Pisaura mirabilis*, where we use experimental data to estimate a key model parameter. 
Our results provide a general framework for understanding why nuptial gift-giving evolves in some systems and not in others, how the evolution of nuptial gift-giving can give rise to sexual conflict, and it provides insight into the evolution of worthless and deceitful nuptial gifts, which occur in several different taxonomic [@Ghislandi2014; @LeBas2005].

Results
================================================================================

Analytical model
--------------------------------------------------------------------------------

We use a time-in and time-out model [@Clutton-Brock1992; @Kokko2001; @Kokko2006] in which choosy (female) and non-choosy (male) individuals spend some period of time within the mating pool searching for a mate (time-in) followed by a period outside the mating pool (time-out).
During time-out, females spend some duration of time ($T_{\mathrm{f}}$) gestating or rearing (hereafter 'processing') offspring. 
We define the number of offspring produced by a female per reproductive cycle as $\lambda$.
Since females enter time-out after mating, this assumption is equivalent to assuming a system with sequential polyandry.
For simplicity, we assume male time to replenish sperm is negligible, but males can spend some duration of time ($T_{\mathrm{m}}$) out of the mating pool searching for nuptial gifts.

@@

Criteria for male search and female choosiness
--------------------------------------------------------------------------------

The probability $G$ that a male succeeds in securing a nuptial gift is defined by,

<!--

$$G = 1 - e^{-\frac{1}{\alpha}T_{m}}.$$
--->

\begin{equation}
G = 1 - e^{-\frac{1}{\alpha}T_{\mathrm{m}}}.
\end{equation}

In Eq. 1, $\alpha$ defines the expected search time before encountering a nuptial gift.
Thus, the probability of finding a nuptial gift is higher the more time $T_{\mathrm{m}}$ is spent searching. 
During time-in, individuals encounter conspecifics at a rate of $R$.
A focal individual will therefore encounter conspecifics of the opposite sex at a rate of $R/2$ if the ratio of males to females in the mating pool ($\beta$) is equal.
More generally, males will be encountered at a rate of $R\beta/(\beta+1)$, and females will be encountered at a rate of $R/(\beta+1)$.
An example of how the structure of the time-in time-out model applies to a system with nuptial gift-giving is given in Figure 1.
We assume that mating with a nuptial gift increases the fitness of each offspring by an increment of $\gamma$.
We proceed to find the thresholds $\gamma_{\mathrm{m}}$ and $\gamma_{\mathrm{f}}$ above which males and females are favoured by selection to search for mates with nuptial gifts and exhibit choosiness for nuptial gifts, respectively.
We show (see Methods) that the initial threshold value of $\gamma$ ($\gamma_{\mathrm{m}}$) necessary for males to increase their fitness by investing time searching for a nuptial gift (time that could otherwise be spent searching for a mate) is,

$$\gamma_{m} > \alpha \frac{M}{\sqrt{\beta}}.$$
--->

\begin{equation}
\gamma_{\mathrm{m}} > \alpha \frac{R}{\beta + 1}.
\end{equation}

Inequality 2 means that if nuptial gifts are not abundant and thus require a long time to find (i.e., high $\alpha$), or if males encounter many females per unit time (i.e., high $R / (1+\beta)$), then the nuptial gift must result in a high fitness increment for selection to favour gift-searching. 
In general, when ineq. 2 is satisfied, we predict selection to favour the evolution of nuptial gift-giving.

We can similarly predict the conditions for which there is selection for female choosiness. 
If $\gamma$ is sufficiently high, then females increase their fitness by rejecting males without gifts and mating only with males that provide nuptial gifts. 
To illustrate, we assume that all males in a population search for a duration of $T_{\mathrm{m}}$, in which case the threshold fitness increment for females ($\gamma_{\mathrm{f}}$) is,

$$\gamma_{f} > \frac{1}{T_{f} M \sqrt{\beta} \left(1 - e^{-\frac{1}{\alpha}T_{m}} \right)}.$$

\begin{equation}
\gamma_{\mathrm{f}} > \frac{1}{T_{\mathrm{f}} R \left(\frac{\beta}{\beta + 1}\right) \left(1 - e^{-\frac{1}{\alpha}T_{\mathrm{m}}} \right)}.
\end{equation}

Inequality 3 shows that as offspring processing time ($T_{\mathrm{f}}$), mate encounter rate ($R\beta / (\beta + 1)$), or the probability of a male finding a nuptial gift ($1 - \exp(-T_{\mathrm{m}}/\alpha)$ ) decreases, the threshold value of fitness above which selection will favour choosiness ($\gamma_{\mathrm{f}}$) increases.
This can be understood intuitively by realising that rejecting a prospective male represents an opportunity cost for the female. 
This opportunity cost becomes small if many males with gifts are encountered, hence the appearance of the rate at which males with gifts are encountered in the denominator.
Figure 2 shows how $\gamma_{\mathrm{m}}$ and $\gamma_{\mathrm{f}}$ change with increasing $\alpha$. 
For $\gamma_{\mathrm{f}}$, we assume that males search for the expected time required to obtain a nuptial gift ($T_{\mathrm{m}} = \alpha$). 
Note that $\beta$ does not have a closed form solution given $R$, $T_{\mathrm{f}}$, and $T_{\mathrm{m}}$, so $\beta$ was calculated using recursion (see Methods).

```{r, echo = FALSE, fig.width = 8, fig.height = 4, fig.cap = "Fitness thresholds above which males increase their fitness by searching for nuptial gifts (blue lines; Eq. 2) and females increase their fitness by rejecting males that do not offer gifts (red lines; Eq. 3). Parameter space includes areas in which males do not search for nuptial gifts and females are not choosy (A), males search but females are not choosy (B), females would be choosy but males do not search (C), and males search and females are choosy (D). Arrows in panel a indicate the effect of increasing interaction rate ($R$), female time-out ($T_{\\mathrm{f}}$), and male search time ($T_{\\mathrm{m}}$). As an example, trajectories for $T_{\\mathrm{f}} = 2$, and $T_{\\mathrm{m}} = \\alpha$ are shown for values of $R=2$ (panel a) and $R = 1$ (panel b). Females are assumed to be the choosy sex, which is maintained as long as $\\alpha < T_{\\mathrm{f}}$."}
alpha_vals <- seq(from = 0.0001, to = 2, by = 0.0001);
par(mar = c(5, 5, 1, 1), mfrow = c(1, 2));
beta_vals <- get_beta(M = 2, TF = 2, TM = alpha_vals);
g_m <- gamma_m(R = 2, beta = beta_vals, alpha = alpha_vals, TF = 2);
g_f <- gamma_f(R = 2, beta = beta_vals, alpha = alpha_vals, TF = 2);
plot(x = alpha_vals, y = g_m, type = "l", lwd = 2, cex.lab = 1.25,
     xlab = expression(paste("Nuptial gift search time (", alpha, ")")),
     ylab = expression(paste("Fitness thresholds (", gamma[f], " & ", gamma[m], 
                             ")")),
     ylim = c(0, 2), col = "black", xaxs="i", yaxs="i");
points(x = alpha_vals, y = g_f, type = "l", lwd = 2, cex.lab = 1.25,
     col = "black", lty = "solid");
g_v <- which(g_f > g_m);
g_w <- which(g_f <= g_m);
polygon(x = c(alpha_vals[g_v], rev(alpha_vals[g_v])), 
        y = c(g_m[g_v], rev(g_f[g_v])),
        border = NA, col = "grey70");
polygon(x = c(alpha_vals[g_w], rev(alpha_vals[g_w])), 
        y = c(g_m[g_w], rev(g_f[g_w])),
        border = NA, col = "grey20");
bott <- rep(0, times = length(alpha_vals));
topp <- c(g_m[g_v], g_f[g_w]);
polygon(x = c(alpha_vals, rev(alpha_vals)),
        y = c(topp, rev(bott)),
        border = NA, col = "grey90");
points(x = alpha_vals, y = g_m, type = "l", lwd = 2, cex.lab = 1.25,
       col = "blue");
points(x = alpha_vals, y = g_f, type = "l", lwd = 2, cex.lab = 1.25,
       col = "red");
text(x = 0.8, y = 0.15, cex = 2, labels = "A");
text(x = 0.25, y = 0.4, cex = 1.5, labels = "B");
text(x = 1.8, y = 1.1, cex = 4, labels = "C", col = "white");
text(x = 0.2, y = 1.75, cex = 5, labels = "D");
arrows(x0 = alpha_vals[10000], x1 = alpha_vals[9000], col = "blue",
       y0 = g_m[10000], y1 = 0.9, length = 0.1, lwd = 2);
text(x = alpha_vals[8300], y = 1.0, labels = "R", col = "blue");
arrows(x0 = alpha_vals[15000], x1 = alpha_vals[15000], col = "red",
       y0 = g_f[15000], y1 = 0.5, length = 0.1, lwd = 2);
text(x = alpha_vals[15000], y = 0.415, labels = "R", col = "red");
arrows(x0 = alpha_vals[12000], x1 = alpha_vals[12000], col = "red",
       y0 = g_f[12000], y1 = 0.5, length = 0.1, lwd = 2);
text(x = alpha_vals[12000], y = 0.415, labels = expression(T[f]), col = "red");
arrows(x0 = alpha_vals[18000], x1 = alpha_vals[18000], col = "red",
       y0 = g_f[18000], y1 = 0.5, length = 0.1, lwd = 2);
text(x = alpha_vals[18000], y = 0.415, labels = expression(T[m]), col = "red");
text(x = 1, y = 1.9, labels = "R = 2", cex = 1.5);
text(x = 1.8, y = 1.915, labels = "a", cex = 1.5, col = "black");

The analytical framework predicts 4 zones, which are delineated by inequalities 2 and 3 and describe the initial thresholds for favouring search of nuptial gifts in males and choosiness for nuptial gifts in females (Figure 2a).
Consequently, the modelling framework gives a description of the conditions under which nuptial gift-giving is expected to occur (Figure 2a, Zone D) and the conditions under which only selection for male searching (Figure 2a, Zone B) or female choosiness (Figure 2a, Zone C) are predicted. 
These results therefore highlight the potential for sexual conflict over nuptial gift-giving.

Evolution of male search and female choosiness
--------------------------------------------------------------------------------

We used an individual-based model (IBM) to simulate the evolution of nuptial gift-giving and female choosiness from an ancestral condition in which neither exists. 
The IBM was written to satisfy the assumptions of our analytical time-in and time-out model as much as practical (see Supporting Information S1). 
Using the IBM, we modelled a spatially-implicit, finite population of females and males. 
At each time step, some individuals enter or remain within the mating pool (time-in), where they potentially interact and mate. 
After mating, males and females may leave the mating pool to search for nuptial gifts and to produce offspring, respectively (time-out). 
Mortality occurs with a fixed probability in each time step, then a ceiling regulation is applied to limit population growth (see Methods).

The rates at which females are encountered by males $R_{\mathrm{f},\mathrm{m}}$ and males with nuptial gifts $R_{\mathrm{m_{G}}, \mathrm{f}}$ are encountered by females are both calculated directly from the IBM, thereby modelling how these rates might be estimated from empirical data, so the male threshold for increasing fitness by searching is,

$$\gamma_{m, IBM} > \alpha M_{m}.$$

\begin{equation}
\gamma_{\mathrm{m, IBM}} > \alpha R_{\mathrm{f},\mathrm{m}}.
\end{equation}

Similarly, female threshold for increasing fitness by choosiness is,

$$\gamma_{f, IBM} > \frac{1}{T_{m}M_{f,G}}.$$

\begin{equation}
\gamma_{\mathrm{f, IBM}} > \frac{1}{T_{\mathrm{f}}R_{\mathrm{m_{G}}, \mathrm{f}}}.
\end{equation}

Consequently, the IBM and the analytical model differ slightly (e.g., time is discrete in the IBM but continuous the analytical model, and in the IBM, a fitness increment is applied to the focal female in the form of birthrate increase rather than offspring fitness; see Supporting Information S1 for details). 
But the predicted thresholds are theoretically equivalent and yield predictions that are qualitatively the same (Figure 3).

```{r echo = FALSE, fig.width = 6, fig.height = 6, fig.cap = "The coevolution of male search and female choosiness as a function of nuptial gift search time ($\\alpha$). Points show where the lower 95% confidence interval of female choosiness (red) and male search (blue) exceeds zero, indicating evolution of choosiness or nuptial gift search. Each point includes data from 3200 replicate simulations with identical starting conditions. Red and blue lines show thresholds above which the mathematical model predicts that females should be choosy and males should search, respectively (in agreement with Figure 2). Minor deviations between the analytical model and the simulation results are expected due to the finite population size, the substantial stochasticity inherent to the simulation model, and the potential for coevolution (for details, see Supporting Information S1). The number of individuals in the population remained at or near carrying capacity of $K = 1000$. In each time step, up to 3000 total pair-wise interactions occurred. Expected female processing time was set to $T_{\\mathrm{f}}=2$ time steps, and $\\gamma$ and $\\alpha$ values in the range [0.0, 2.0] and [0.1, 2.0], respectively, were used."}
SUMM         <- read.csv(file = "inst/summary_coevolving.csv")
alpha_vals   <- sort(unique(SUMM[,1]));  
TM_vals      <- tapply(X = SUMM[,3],  INDEX = SUMM[,1], FUN = mean);
m_vals       <- tapply(X = SUMM[,13], INDEX = SUMM[,1], FUN = mean);
beta_vals    <- tapply(X = SUMM[,14], INDEX = SUMM[,1], FUN = mean);
Male_M       <- tapply(X = SUMM[,15], INDEX = SUMM[,1], FUN = mean);
Female_M     <- tapply(X = SUMM[,16], INDEX = SUMM[,1], FUN = mean);
Mf_nupt      <- tapply(X = SUMM[,17], INDEX = SUMM[,1], FUN = mean);
Mf_nonupt    <- tapply(X = SUMM[,18], INDEX = SUMM[,1], FUN = mean);

par(mar = c(5, 5, 1, 1)); 
plot(x = alpha_vals, y = g_f, type = "l", lwd = 2, cex.lab = 1.25, col = "red",
     xlab = expression(paste("Nuptial gift search time (", alpha, ")")),
     ylab = expression(paste("Fitness threshold (", gamma[f], " & ",  gamma[m], 
                             ")")),
     ylim = c(0, 2));

points(x = alpha_vals, y = g_m, type = "l", lwd = 2, col = "blue");

for(i in 1:dim(SUMM)[1]){
  if(SUMM[i, 7] > 0 & SUMM[i, 8] > 0){
    points(x = SUMM[i, 1], y = SUMM[i, 2], cex = 2, pch = 20, col = "red");
  }
}
```

>>>>

---
title: "A general model for the evolution of nuptial gift-giving"
author: |
  | Anders P. Charmouh$^{1,4, 5}$, Trine Bilde$^{2}$, Greta Bocedi$^{1}$, A. Bradley Duthie$^{3}$
date: '[1] School of Biological Sciences, University of Aberdeen, UK. [2] Department of Biology, Aarhus University, Denmark. [3] Biological and Environmental Sciences, University of Stirling, UK.  [4] Bioinformatics Research Centre Aarhus University, University City 81, building 1872, 3rd floor, DK-8000 Aarhus C, Denmark [5] Email: apc@birc.au.dk'
output:
  pdf_document:
    fig_caption: yes
    keep_tex: yes
  html_document: default
  word_document:
    fig_caption: yes
    pandoc_args:
    - --csl
    - evolution-letters.csl
    reference_docx: docx_template.docx
bibliography: nuptial_refs.bib
header-includes:
- \usepackage{amsmath}
- \usepackage{natbib}
- \usepackage{lineno}
- \usepackage{caption}
- \usepackage[utf8]{inputenc}
- \bibliographystyle{amnatnat}
linestretch: 1
link-citations: yes
linkcolor: blue
csl: evolution-letters.csl
biblio-style: apalike
---

```{r, echo = FALSE}
library("knitr")
calc_beta <- function(M = 1, TF = 1, TM = 1){
  term1 <-   (M * (TF - TM))^2;
  term2 <-   (M * (TF - TM))^4;
  term3 <- 4*(M * (TF - TM))^2;
  ttrm  <- (term2 + term3)^(1/2);
  sol1  <- 0.5 * (term1 - ttrm) + 1;
  sol2  <- 0.5 * (term1 + ttrm) + 1;
  if(TF > TM){
    beta = max(c(sol1, sol2));
  }else{
    beta = min(c(sol1, sol2));
  }
  return(list = c(beta = beta, solution1 = sol1, solution2 = sol2));
}
get_beta <- function(M = 1, TF = 1, TM = 1){
  betas <- rep(NA, length = length(TM));
  for(i in 1:length(TM)){
    betas[i] <- calc_beta(M = M, TF = TF, TM = TM[i])[[1]];
  }
  return(betas);
}
PrGf <- function(Tm, a1){
  vals <- 1 - exp(-(1/a1)*Tm);
  return(vals);
}
PrFf <- function(Tm, a1, a2 = Inf){
  vals <- (1 - exp(-(1/a2)*Tm)) * exp(-(1/a1)*Tm);
  return(vals);
}
dWdT <- function(gamma, Tm, beta = 1, M = 100, a1 = 1){
  num1 <- gamma * (M*M + ((M*M*Tm + M*sqrt(beta))/a1) );
  dem1 <- exp((1/a1)*Tm);
  num2 <- (num1 / dem1) - M*M*(1 + gamma);
  den2 <- ((Tm * M) + sqrt(beta))^2;
  val  <- num2/den2;
  return(val);
}
dWdT_check <- function(gamma, beta = 1, M = 100, a1 = 1){
  term1 <- (gamma * M) / (a1*sqrt(beta));
  term2 <- (M*M) / beta;
  val   <- term1 - term2;
  return(val);
}
recursive_b <- function(B, D, Tf, Tm, crit = 0.0001, maxit = 9999){
  conv <- 1;
  iter <- 0;
  while(conv > crit & iter < maxit){
    Fe   <- D * (B / (1 + B));
    Me   <- (1 + B) / D;
    Bn   <- (Me / (Tm + Me)) / (Fe / (Tf + Fe));
    iter <- iter + 1;
    conv <- abs(Bn - B);
    B    <- Bn;
  }
  return(list(B = B, conv = conv, iter = iter));
}
calc_beta <- function(M = 1, TF = 1, TM = 1){
  vals <- recursive_b(B = 1, D = M, Tf = TF, Tm = TM);
  beta <- vals$B
  sol1 <- vals$conv;
  sol2 <- vals$iter;
  return(list = c(beta = beta, solution1 = sol1, solution2 = sol2));
}
get_beta <- function(M = 1, TF = 1, TM = 1){
  betas <- rep(NA, length = length(TM));
  for(i in 1:length(TM)){
    betas[i] <- calc_beta(M = M, TF = TF, TM = TM[i])[[1]];
  }
  return(betas);
}
gamma_m <- function(R = 1, beta = 1, alpha = 1/16, TF = 2){
  return(alpha * (R / (beta + 1)));
}
gamma_f <- function(R = 1, beta = 1, alpha = 1/16, TF = 2){
  topp <- 1;
  bott <- TF * R * (beta / (beta + 1)) * (1 - exp(-alpha/alpha));
  return(topp/bott);
}
PrGf <- function(Tm, a1){
  vals <- 1 - exp(-(1/a1)*Tm);
  return(vals);
}

@@

Nuptial gift-giving occurs in several taxonomic groups including insects, snails, birds, squid, arachnids and humans. 
Although this trait has evolved many times independently, no general framework has been developed to predict the conditions necessary for nuptial gift-giving to evolve. 
We use a time-in time-out model to derive analytical results describing the requirements necessary for selection to favour nuptial gift-giving. 
Specifically, selection will favour nuptial gift-giving if the fitness increase caused by gift-giving exceeds the product of expected gift search time and encounter rate of the opposite sex. 
Selection will favour choosiness in the opposite sex if the value of a nuptial gift exceeds the inverse of the time taken to produce offspring multiplied by the rate at which mates with nuptial gifts are encountered. 
Selection can differ between the sexes, potentially causing sexual conflict. 
We further investigate these results using an individual-based model inspired by a system of nuptial gift-giving spiders, *Pisaura mirabilis*, by estimating the fitness benefit of nuptial gift-giving using experimental data from several studies. 
Our results provide a general framework for understanding when the evolution of nuptial gift-giving can occur and provide novel insight into the evolution of worthless nuptial gifts, occurring in multiple taxonomic groups with implications for understanding parental investment.

@@

Nuptial gift-giving occurs when the choosy sex (usually the female) receives gifts from the opposite sex (usually the male) during courtship. 
It is a widespread phenomenon, occurring within several diverse taxonomic groups such as insects, snails, birds, squid, arachnids and humans [@Lewis2014; @Lewis2012; @Albo2014]. 
Despite the ubiquity of this behaviour, little effort has been made to conceptualise the evolution of nuptial gift-giving within a general modelling framework [@Lewis2014; @Iwasa2022]. 
Recent models describing the evolution of nuptial gift-giving have focused on co-evolution between male nuptial gift-giving and female propensity to remate, and evolutionarily stable nuptial gift sizes [@Iwasa2022; @Kamimura2021], but a general framework describing the conditions under which nuptial gift-giving can be initially favoured by selection is needed to understand when gift-giving should evolve.

Nuptial gift-giving may allow males to increase fitness by acquiring additional mates, indirect benefits (by increasing offspring fitness), prolonged copulations, and success in sperm competition [@Lewis2014; @Albo2013; @Ghislandi2014]. 
However, this potential fitness increase comes at the expense of producing a nuptial gift, which may be costly in terms of time and resources. 
Females may increase their fitness by receiving nutritionally valuable nuptial gifts, but expressing a preference for males with gifts might result in a mating opportunity cost if available males without gifts are rejected. 
With respect to nuptial gift-giving, the evolutionary interests of both sexes may not always fully overlap. 
This can cause sexual conflict, which is a difference in the fitness interests between sexes that occurs when an interaction results in a situation where individuals cannot both achieve an optimal outcome [@Parker2006].
For example, under some conditions, it might be optimal for males but not females to mate if females do not benefit from mating with males without nuptial gifts.

Much work has sought to explain how gift-giving tactics are maintained, with explanations including condition-dependent strategies, gift-giving as a way to decrease female aggression during copulation, or gifts as sensory traps [@Albo2019; @Lubin2007; @Ghislandi2018; @Toft2016]. 
An example of such a system is the nuptial gift-giving nursery-web spider *Pisaura mirabilis*, where males may court females with or without nuptial gifts [@Bristowe1926; @Tuni2013a]. 
Here, males may provide females with costly nuptial gifts in the form of captured arthropod prey, and females may exhibit preference for males with a nuptial gift by rejecting males without a nuptial gift [@Albo2013].

We develop a general framework for investigating the evolution of nuptial gift-giving and choosiness using a time-in, time-out modelling approach and an individual-based model [@Clutton-Brock1992]. 
Specifically, we derive conditions under which selection will favour male search for nuptial gifts and female rejection of gift-less males. 
We show that selection for searching and choosiness depends on whether a threshold fitness value of the nuptial gift is exceeded. 
Our model demonstrates the importance of nuptial gift cost, sex ratio, and mate encounter rate in determining the threshold above which selection will favour the evolution of nuptial gift-giving.
Importantly, we show that the threshold value differs for male searching and female choosiness. 
We complement the predictions of our analytical model by formulating an individual-based model, which further supports the main theoretical results. 
We apply our model to an example system with nuptial gifts, the nursery web spider *Pisaura mirabilis*, where we use experimental data to estimate a key model parameter. 
Our results provide a general framework for understanding why nuptial gift-giving evolves in some systems and not in others, how the evolution of nuptial gift-giving can give rise to sexual conflict, and it provides insight into the evolution of worthless and deceitful nuptial gifts, which occur in several different taxonomic [@Ghislandi2014; @LeBas2005].

Model
================================================================================

We use a time-in and time-out model [@Clutton-Brock1992; @Kokko2001; @Kokko2006] in which choosy (female) and non-choosy (male) individuals spend some period of time within the mating pool searching for a mate (time-in) followed by a period outside the mating pool (time-out).
During time-out, females spend some duration of time ($T_{\mathrm{f}}$) gestating or rearing (hereafter 'processing') offspring. 
We define the number of offspring produced by a female per reproductive cycle as $\lambda$.
Since females enter time-out after mating, this is equivalent to assuming a system with sequential polyandry.
For simplicity, we assume male time to replenish sperm is negligible, but males can spend some duration of time out of the mating pool searching for nuptial gifts.
We define $T_{\mathrm{m}}$ as the time until a gift is found by a searching male.

@@

Analytical model
-------------------------------------------------------------------------------

The probability that a male succeeds in securing a nuptial gift ($G$) after a duration of $T_{\mathrm{m}}$ is defined by,

$$P(G) = 1 - e^{-\frac{1}{\alpha}T_{m}}.
\tag{1}
$$

In Eq. 1, $\alpha$ defines expected search time before encountering a nuptial gift.
Thus, the probability of finding a nuptial gift is higher the more time is spent searching. 
During time-in, individuals encounter conspecifics at a rate of $R$.
A focal individual will therefore encounter conspecifics of the opposite sex at a rate of $R/2$ if the ratio of males to females in the mating pool ($\beta$) is equal.
More generally, males will be encountered at a rate of $R\beta/(\beta+1)$, and females will be encountered at a rate of $R/(\beta+1)$.
An example of how the structure of the time-in time-out model applies to a system with nuptial gift-giving is given in Figure 1.
We assume that mating with a nuptial gift increases expected offspring production by an increment of $\gamma$.
Mortality occurs for females and males in ($\mu_{\mathrm{i,f}}$, $\mu_{\mathrm{i,m}}$) and out ($\mu_{\mathrm{o,f}}$, $\mu_{\mathrm{o,m}}$) of the mating pool. 
Following @Kokko2006, we assume $m_{\mathrm{i,f}} = m_{\mathrm{o,f}} = 1$ and $m_{\mathrm{i,m}} = m_{\mathrm{o,m}} = 1$.

First, we find the fitness of males that search versus do not search for a nuptial gift. 
We then describe the fitness consequences of female choice to accept or reject males based on their provision of a nuptial gift.
We can then calculate the thresholds $\gamma_{\mathrm{m}}$ and $\gamma_{\mathrm{f}}$ above which males and females are favoured by selection to search for nuptial gifts and exhibit choosiness for nuptial gifts, respectively.

*Male fitness*

During time-out, males can search for a nuptial gift.
Males can adopt one of two strategies; either search or do not search for a nuptial gift. 
Males with the searching strategy continue to search until they find a nuptial gift, while males that do not search immediately re-enter the mating pool. 
Consequently, time searching for a nuptial gift will come at the cost of mating opportunities but might increase offspring fitness. 
We therefore need to model the expected length of time spent outside of the mating pool for males that search for nuptial gifts ($E[T_{\mathrm{m}}]$), which equals $\alpha$. 
We can integrate search time $t$ over the rate at which nuptial gifts are encountered ($\exp(-1/\alpha)$) to show $E[T_{\mathrm{m}}] = \alpha$,
